Indigenous Subsistence and Colonial Encounters: A Zooarchaeological Perspective from Kiwulan, a late Iron site in Northeastern Taiwan

### 2025-03-05 analysis/paper/paper.qmd
work on discussion, conclusion, and abstract sections

---
title: "Exploring colonial impacts on indigenous hunting and animal use in northeastern Taiwan"
author:
  - Li-Ying Wang:
      correspondence: "yes"
      email: liyingw15@gmail.com
      orcid: 0000-0001-9060-2125
      institute: Academia Sinica
title-block-published: "Last updated"  
date: now
date-format: long
format: 
  docx:
    reference-doc: "../templates/template.docx" # Insert path for the DOCX file
execute:
  echo: true
  warning: false
  message: false
  comment: "#>"
  fig-path: "../figures/"
  fig-dpi: 600
filters:
  - ../templates/scholarly-metadata.lua
  - ../templates/author-info-blocks.lua
  - ../templates/pagebreak.lua
bibliography: references.bib
csl: "../templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
abstract: |
  Text of abstract
keywords: |
  keyword 1; keyword 2; keyword 3
highlights: |
  These are the highlights. 
---

@@

# Discussion
<!--results and chronology -->
The sampled faunal assemblage from the major residential area at Kiwulan suggests a continuation of sika deer consumption from CL6 to CL2, and then shifted to cattle in CL1. Despite deer was a substantial part of dietary practices at Kiwulan, variations in their quantity, ubiquity, butchering patterns, and proportions relative to other animals show changes that could reflect the indigenous reactions to foreign presence. Based on radiocarbon dates and diagnostic artifacts indicating specific time [@Wang2020ornament; @Chen2007], the cultural layers were further grouped into temporal sequences in relation to foreign presences for further discussion. The temporal sequences include the pre-European period (CL6, CL5), the European period (CL4), the post-European period (CL3), and the arrival of Han-Chinese immigrants (CL2, CL1). It should be noted that the European period does not necessarily mean a solid European colonial control over the Kiwulan, but a broad impact from European occupation in Northern Taiwan and associated colonial activities on indigenous societies in Yilan, such as getting involved in the long-distance trade network, being part of the tribute system, and experiencing occasional European attacks.

<!--summary of overall observations across time periods -->
Faunal remains recovered from the European period reveal the highest quantity, followed by the late pre-European period. During the European period, in addition to the predominance of deer, an increased proportion of suid suggests a slightly greater reliance on suid as a secondary meat source. A similar pattern was observed from a higher ubiquity of suids during the European period compared to other periods. Interestingly, deer bones from the late pre-European period presented the same ubiquity as in the European period but in only half the quantity. This might indicate consistency in the frequency and distribution of deer at the settlement scale before and after European presence, but deer utilization was intensified during the European period. The dramatic decrease in deer bones after the European period might be related to both the decline of the Kiwulan population and a decrease in the wild deer population. Sika deer populations in western Taiwan have significantly shrunk since the 18th century because of habitat loss and overhunting [@Chen2000animal]. Despite this, deer remained a predominant resource for local consumption at Kiwulan during the post-European and early Chinese period. In the late Chinese period, the dietary preference shifts from deer to cattle, suggesting a Chinese influence and the relocation of the Kavalan people to Hualien due to population pressures from Chinese immigration [@Zhan1995].

<!--skeletal part frequencies and butchery patterns of deer-->
Despite being the main resource throughout most periods, deer utilization shows different patterns over time. Dear body parts and butchery patterns demonstrate more complete carcasses processed within the settlement during the European and pre-European period compared to the post-European and Chinese period. Although more complete and evenly distributed joints with cut marks were found in both the European period and before the European arrival, the substantial rise in head parts (e.g., crania and mandibles) with the overall high quantity suggests more deer were brought back intact to the settlement. On the contrary, the low frequency of head parts in the early pre-European period and Chinese period might indicate deer could be mostly processed at the hunting location with mostly meaty (e.g., upper limb) and mid-meaty portions (e.g., ribs) selectively butchered and transported back. Interestingly, among the meaty portions, front quarters (e.g., humeri, radius) are more frequently found compared to forequarters (e.g., femur, tibia) throughout all periods. One likely explanation is that femur and tibia might be heavily fragmented due to being broken for their high marrow yields [@Madrigal2002], leading to underestimation due to difficult identification. Marrow extraction at Kiwulan can be supported by around one-third of long bones being fractured and a moderate degree of fragmentation on average. Another possibility is forequarters could be traded since deer legs and meats as commodities were commonly mentioned in European records [@Blusse2006, pp.220-224], but there is no mention of any specific legs commonly traded. Considering the generally high frequencies of meaty portions found at Kiwulan, deer might not just for consumption at the site, but also for trading.

<!--deer hides-->
Deer hides were important commodities since the mid-15th century, connecting Han Chinese, Japanese, and Indigenous people, and later monopolized by the Europeans in the 17th century, when deer products were largely exported to Japan, especially from western Taiwan [@Cheng2017; @Hollmann1991]. The cut marks found on the lower limbs of deer at Kiwulan should be associated with the traces left by skinning, which confirms the removalof  deer skins and likely being traded with external traders. A large quantity of foreign goods, such as ornaments, porcelains, and metal items, supports frequent exchange between Kiwulan residents and outsiders [@Wang2020ornament; @Chen2007]. The evidence of skinning and foreign goods at Kiwulan suggests that not only western Taiwan but also northeastern Taiwan was involved and impacted by the long-distance trade. Although deer bones with skinning cuts did not show a substantial increase in the European period, the disproportionate increase in head parts might hint at a likelihood of brain tanning when preparing hides. Brain tanning is a traditional method for softening and preserving hides commonly used in many Indigenous cultures, such as North America, Southern Africa, North Asia, and Japan [@Baillargeon2010; @Badenhorst2009; @Hatt1969; @Matsumoto2016]. The present-day Bunun group in the central highlands of Taiwan has revived hide tanning based on oral history, primarily using peanuts or pig lard, with a few examples using animal brains [@Li1999, pp. 48; @Xu1994]. Despite the materials for tanning used by the Kavalan people being unknown due to the loss of traditions, brain tanning is a practicable method given the availability of deer brains from hunting.

<!-- population dynamics and arguments-->
The population dynamics at Kiwulan must be considered when discussing faunal changes over time. A larger pottery vessels during the European period suggests population growth [@Wang2020pot; @Hsieh2009], which may contribute to the increasing animal consumption overall. However, the impacts from foreign presence on faunal assemblages can be revealed through butchery patterns of deer parts, the relative proportions of animals, and taxon ubiquity. First, deer bones with skinning evidence were mostly found during the European period, followed by the late pre-European phase, corresponding with the rise of regional trade networks since the 15th century and intensified in the 17th century. Second, if the high frequency of deer head parts during the European period solely resulted from food scarcity due to population growth and increased consumption of brains for nutrition, a parallel rise in marrow extraction would be expected. However, fractured long bones did not increase during the European period or the previous phase; instead, they were 10-30 % lower than in other periods. Third, population growth should have led to a more proportional increase in both large and small animals. Although there was a high taxonomic diversity during the European period, suids present a disproportionate rise in their ubiquity, quantity, and proportion compared to other animals. This suggests that intensified deer hunting for trade may have driven indigenous people to rely more on suids as an alternative prey, likely due to their comparable body size and meat return rates to deer [@Broughton2011prey]. Suids as a secondary protein source after deer is also observed in prehistoric western Taiwan [@Chuang2021].

<!--indigenous reaction and resilience-->
Compelling the results of faunal analysis and reports from historical records, the substantial increase in deer bones in the European period not just simply the result of population growth, but also a shift driven by European presence and associated intensified long-distance trade. It is clear to see deer is both the first priority of food source and commodities at Kiwulan since the late pre-European period and intensified during the European period. Based on the dietary preference, the increased use of suids during the European period could be a temporary subsistence strategy in response to the high demands of deer products. After the Europeans left, deer hide trade continued but on a smaller scale, with around 10,000 hides exported annually in the 18th century, which is 20 times less than during the European period [@]. This change likely reduced the reliance of suids, as shown by their decreased proportion at Kiwulan. The slightly increase in muntjact during the post-European and early Chinese period might be associated with the decline of deer population, leading people turned to use muntjact hides for trade instead []. Nevertheless, deer continued to be consumed until the early Chinese period. Everyday practices play an essential role for connecting history and social memories, enabling Indigenous people maintain tradition in a setting of frequent cultural exchange or asymmetrical relations [@Silliman2009; e.g., @Noe2023]. The persistence of deer consumption at Kiwulan reflect identity negotiation and tradition maintenance, while the increased utilization of suids likely responded to the high demand for deer products in large scale trade. This highlights how Kiwulan residents reacted to foreign presence and demonstrated cultural resilience.

<!--taphonomic & preservation bias -->

# Conclusion

Through contextralizing the faunal remains into time periods relaitve to different foreign presences, including the Europeans in the 17th century corresponding to CL4, and the Chinese immigrants since the late 19th century corresponding to CL1, the resutls reveals how indigenous people respond to the foreign influences on their economic aspect.

By situating the findings within a pericolonial context (Acabado, 2017), locally made pottery and faunal remains reveal cultural persistence and resilience among Indigenous peoples in northeast Taiwan. The interaction between Europeans and Indigenous peoples in the region appears to have been based on negotiation and bartering (Torrence, 2000). For instance, the Dutch reduced and eventually waived annual tribute in 1648 in response to the resistance to paying tribute by most Indigenous settlements in Yilan (Blussé and Everts, 2000, pp. 534, 536, 558; Kang, 2012). The continuous consumption of deer and consistent skinning patterns before and after the arrival of Europeans indicate a more adaptable tribute payment system, whereby the Indigenous peoples did not need to hunt more deer to meet European demand. However, it is important to note that there may have been a gradual decline in deer numbers since the post-European period, which is supported by the increased number of pigs found in the same period. This could suggest cultural resilience through dietary changes to adapt to deer overhunting. A more comprehensive investigation across different Indigenous settlements in northeastern Taiwan could provide a clearer understanding of overall dietary patterns.

Taiwan has played an important role in exploring the effect of colonial activities on Indigenous communities and local environment since the protohistoric period. This project will explore the environmental effect of colonial presence on local ecology on a regional scale using archaeological evidence that has not yet been fully understood. Archaeological materials can offer more direct evidence, allowing better interpretation for the environments where Indigenous live in that can also reduce biases from historical records. Through exploring the long-term dynamics of human-environmental relationship in colonial settings, we can better understand changes in local environments and Indigenous resilience. In addition to the Europeans, the interaction between Chinese traders and Indigenous groups in Taiwan since the Neolithic period, and an influx of Han Chinese immigrants in the 18th century, has contributed to the significant local cultural and environmental transformation. Those historical events are closely related to the Taiwan Indigenous people today, which is crucial to understanding the contemporary Indigenous cultures and socio-ecological landscape.

The results of this project can advance our knowledge of the reaction of Indigenous people in Taiwan when facing dominant foreign power, and generate new anthropological insights about the adaptation and choices of Indigenous people. In the long term, this research will also benefit local Indigenous people, Kavalan, in northeastern Taiwan and help them to trace back their past and revitalize their cultural heritage by the plans of public presentation in the future. In addition, this project will combine geoarchaeological, geochemical, and computational methods to analyze archaeological evidence. The multidisciplinary approach can not only advance the development and application of scientific techniques in Taiwan archaeology, but also provide an opportunity to form collaborative relationships with scholars with different backgrounds, which will help to develop the academic network and create interdisciplinary collaboration.

This research will contribute to a broader understanding of the diverse ways Indigenous communities adapted to and resisted colonial forces, enhancing our comprehension of the long-term impacts of European colonization on local societies.

>>>>

---
title: "Indigenous Subsistence and Colonial Encounters: A Zooarchaeological Perspective from Kiwulan, a late Iron site in Northeastern Taiwan"
author:
  - Li-Ying Wang:
      correspondence: "yes"
      email: liyingw15@gmail.com
      orcid: 0000-0001-9060-2125
      institute: Academia Sinica
title-block-published: "Last updated"
date: now
date-format: long
format: 
  docx:
    reference-doc: "../templates/template.docx" # Insert path for the DOCX file
execute:
  echo: true
  warning: false
  message: false
  comment: "#>"
  fig-path: "../figures/"
  fig-dpi: 600
filters:
  - ../templates/scholarly-metadata.lua
  - ../templates/author-info-blocks.lua
  - ../templates/pagebreak.lua
bibliography: references.bib
csl: "../templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
abstract: |
  Colonial encounters between Europeans and Indigenous people often led to profound impacts on many Indigenous societies and cultures around the world. In northeastern Taiwan, local Indigenous groups experienced foreign presences through regional trade networks in East Asia, which later expanded into a larger-scale trade network after the arrival of Europeans in the 17th century. This research focuses on local subsistence through faunal remains to understand how Indigenous communities responded to the regional trade activities driven by the European and Han Chinese, particularly the high demand for natural resources, such as deer. By analyzing faunal remains from the upper cultural layer of Kiwulan (AD 1450-1950), a late Iron Age site in northeastern Taiwan, the local participation in deer hide trade and their adaptive strategies were discussed. The results suggest indigenous cultural persistence of deer consumption practiced by local people, along with increased reliance on suids during the European period as the demands for deer products intensified. This paper also provides insights into colonial experiences from Indigenous settlements at the periphery of European territories, a perspective rarely discussed.

keywords: |
  Zooarchaeology; Iron Age; Taiwan; sika deer; trade; European colonization
highlights: |
  These are the highlights. 
---

@@

# Discussion
<!--results and chronology -->
The sampled faunal assemblage from the major residential area at Kiwulan suggests a continuation of sika deer consumption from CL6 to CL2, and then shifted to cattle in CL1. Despite deer was a substantial part of dietary practices at Kiwulan, variations in their quantity, ubiquity, butchering patterns, and proportions relative to other animals show changes that could reflect the indigenous reactions to foreign presence. Based on radiocarbon dates and diagnostic artifacts indicating specific time [@Wang2020ornament; @Chen2007], the cultural layers were further grouped into temporal sequences in relation to foreign presences for further discussion. The temporal sequences include the pre-European period (CL6, CL5), the European period (CL4), the post-European period (CL3), and the arrival of Han-Chinese immigrants (CL2, CL1). It should be noted that the European period does not necessarily mean a solid European colonial control over Kiwulan, but a broad impact from European occupation in Northern Taiwan and associated colonial economic activities on indigenous people in Yilan, such as getting involved in the long-distance trade network, tribute paying, and suffered from occasional European attacks.

<!--summary of overall observations across time periods -->
The sampled area at Kiwulan shows a large quantity of faunal remains in the European period, followed by the late pre-European period. Interestingly, deer bones from the late pre-European period presented the same ubiquity as in the European period but in only half the quantity. This might indicate consistency in the frequency and distribution of deer at the settlement scale before and after European presence, but deer utilization was intensified during the European period. During the European period, in addition to the predominance of deer, an increased relative proportion and ubiquity of suid suggests a greater reliance on suid than in other periods. The dramatic decrease in deer bones after the European period might be related to both the decline of the Kiwulan population and a decrease in the wild deer population. Sika deer populations in western Taiwan have significantly shrunk since the 18th century because of habitat loss and overhunting [@Chen2000animal]. An increase in muntjac use may suggest they served as a substitute, replacing the role of suids in earlier periods. Despite this, deer remained a predominant resource for local consumption at Kiwulan during the post-European and early Chinese period. In the late Chinese period, the dietary preference shifts from deer to cattle, suggesting a Chinese influence and the relocation of the Kavalan people to Hualien due to population pressures from Chinese immigration [@Zhan1995].

<!--skeletal part frequencies and butchery patterns of deer-->
Although deer was the primary resource at Kiwulan, differences in skeletal completeness across periods suggests changes in butchery locations and cooking practices. There were more complete carcasses in the European and late pre-European period compared to the post-European and Chinese periods. Moreover, the substantial rise in fragmented head parts (cranium and mandible) during the European period indicates greater use of soft tissues, such as brains and tongues, or the transport of more intact deer to the settlement for processing. This aligns with the historical account in 1603, Record of Formosa (Dongfan ji) [@Chen1603], which describes deer tongue consumption and group hunting for deer in western Taiwan. Changes in cooking practices are also evident. The decline in burned bones over time with a lowest percentage in the European period suggests a shift from roasting to other cooking methods, such as boiling. The presence of small chopped bones that could fit in local ceramics at Kiwulan indicates boiling as a possible practice since the European period. The shift in cooking practices suggests that the increased need for deer may have led people to adopt boiling due to its additional benefits, such as maximizing residual meat use, facilitating grease rendering, and improving quality of lean meat [@Lupo1997]. This could be further confirmed through organic residue analysis in the future.

A consistent pattern was also identified throughout all periods: High frequencies of upper limb bones but low frequencies of vertebrae and ribs, suggesting selectively butchered of meaty portions. Moreover, it is interesting that front quarters (humerus and radius) are much more abundant than hindquarters (femur and tibia). One possible explanation is that femurs and tibiae were heavily fragmented for marrow extraction due to their high yields [@Madrigal2002; @Lupo1998], making them less identifiable in the assemblage. Marrow extraction can be supported by at least 25% of long bones showing fractures and a moderate degree of fragmentation. Another likelihood is that some hindquarters might be traded, as European records mention deer legs and meats as commodities in addition to hides [@Blusse2006, pp.220-224], but without specifying which legs. Considering the overall high proportion of meaty portions at Kiwulan, deer were likely processed not only for local consumption but also for trade.

<!--deer hides-->
Deer hides trade had prevailed since 15th century in East Asia, and monopolized by the Europeans in the 17th century, when large hides were exported to Japan, especially from western Taiwan [@Cheng2017; @Hollmann1991; @Wang1981]. The cut marks identified on deer lower limbs at Kiwulan confirm the removal of deer skins and likely being traded with external traders. Abundant foreign goods at Kiwulan, including ornaments, porcelains, and metal items, supports frequent exchange between residents and outsiders [@Wang2020ornament; @Chen2007]. This indicates that not only western Taiwan but also northeastern Taiwan was intensively involved and impacted by the long-distance trade. Although lower limbs with skinning cuts did not show a substantial increase in the European period, the disproportionate increase in head parts might hint at a likelihood of brain tanning when preparing hides in addition to tissue consumption. Brain tanning is a traditional way for softening and preserving hides in many Indigenous cultures, such as North America, Southern Africa, North Asia, and Japan [@Baillargeon2010; @Badenhorst2009; @Hatt1969; @Matsumoto2016]. The present-day Bunun group in the central highlands of Taiwan has revived hide tanning based on oral history, primarily using peanuts or pig lard, with a few examples using animal brains [@Li1999, pp. 48; @Xu1994]. Despite the materials for tanning used by the Kavalan people being unknown due to the loss of traditions, brain tanning is a practicable method given the availability of deer heads from hunting.

<!-- population dynamics and arguments-->
The population dynamics at Kiwulan must be carefully considered when interpreting the faunal results, because high population during the European period inferred by pottery could contribute to the increasing animal use overall [@Wang2020pot; @Hsieh2009]. However, the impacts from foreign presence can be still identified through butchery patterns of deer parts, the relative proportions of animals, and taxon ubiquity. First, deer bones with skinning evidence were mostly found during the European period, followed by the late pre-European phase, corresponding with the rise of regional trade networks since the 15th century and intensified trade in the 17th century. Second, if the high frequency of deer parts during the European period, especially heads relative to brains or tongues consumption, solely resulted from food stress due to population growth, a parallel rise in marrow extraction would be expected. In fact, long bones with fracture patterns were 10-30 % lower during the European period  and the previous phase than in other periods. Third, population growth should have led to a more proportional increase in both large and small animals. Although there was a high taxonomic diversity during the European period, suids present a disproportionate rise in their ubiquity, quantity, and proportion. This indicates that trade of deer products may have driven indigenous people to rely more on suids as an alternative prey, likely due to their comparable body size and meat return rates to deer [@Broughton2011prey]. Suids as a secondary protein source after deer are also observed in prehistoric western Taiwan [@Chuang2021]. This hints the role of suids as a supplement when deer was not sufficient.

<!--indigenous reaction and resilience-->
Compelling the zooarchaeological evidence and descriptions from historical records, the substantial increase in deer bones in the European period not just simply the result of population growth, but also reflecting intensified hunting driven by deer trade. The results suggest deer is the first priority of food source and commodities at Kiwulan since the late pre-European period and their economic benefits were intensified during the European period. It should be noted that the role of the Dutch is both colonist who demanded tribute from Yilan, and another trader who were incorporated into the pre-existing inter-insular trade network involving Han Chinese merchants and other indigenous groups in Northern Taiwan [e.g., @Torrence2000]. The prestige values embedded in foreign goods in indigenous culture might lead Kiwulan aggrandizers actively participated in the trade rather than resisted [@Wang2020ornament]. After the Dutch left, deer hide trade continued but back to a small scale, with around 10,000 hides exported annually, which is 10 times less than during the European period [@Nakamura1997]. A small deer trade market and low population at Kiwulan might lead to a reduced reliance of suids, as shown by a low proportion of suids in the post-European period and eventually disappeared during the Chinese period.

<!--Final remarks -->
Everyday practices play an essential role for maintaining Indigenous traditions within a colonial setting or frequent cultural exchange [@Silliman2009; e.g., @Noe2023]. The persistence of deer consumption with similar butchery patterns at Kiwulan reflect a maintained tradition, showing community identities.

while the increased utilization of suids likely responded to the high demand for deer products in large scale trade.  The small sample size after separating them into different periods might limits a further interpretation; however, changes in different phases were still provide evident patterns reflecting changes in animal use. This highlights how Kiwulan residents reacted to foreign presence and demonstrated cultural resilience.

Deer continued to be consumed until the early Chinese period.

# Conclusion
By contextualizing zooarchaeological evidence into time periods relative to foreign presence, including the Europeans in the 17th century, and Chinese immigrants from the late 19th century, this study reveals the participation of indigenous groups in regional trade work. The frequencies, butchering patterns, and human modifications on deer bones confirms the active deer hides trade since the early 15th century and intensified during the European period. The results demonstrates even indigenous people in peripheral areas with relatively weak colonial control were also profoundly impacted than previously thought. In pericolonial setting, the continued prioritization of deer consumption at Kiwulan suggests a cultural persistence of food practices, while the strategic reliance on suids during the European period demonstrates adaptation to high demands for deer products, and later incorporating more muntjac in response to overhunting of sika deer. The long-term dynamics of animal use in northeastern Taiwan can enhance our understanding of Indigenous resilience in face of foreign impacts and environmental change. In addition, the role of indigenous people might be more active in the exchange network considering the large amount of foreign goods that were incorporated into local cultures and used as prestige items. A more comprehensive investigation across Indigenous settlements in northeastern Taiwan and including other dietary evidence, such as botanical remains and organic residue analysis in the future could provide a deeper insights into the regional dietary patterns and human-environment dynamics.
